### 2017-08-28 sec-conclusion.tex
Address copy-editing feedback.

We showed that such a model does, in fact, capture meaningful behavior
patterns and produces descriptions of these behavior patterns that are easy
to interpret. We argued that it is important to capture student behavior
patterns with more sophisticated models than simple discrete distributions
over actions in order to capture information present in bigrams of actions
(or above). By varying the number of latent states inferred, we showed that
the model is flexible and can capture patterns of differing levels of
specificity in this way.  Finally, we investigated whether we can detect
differences in student behavior patterns as they correlate with course
performance. Specifically, we demonstrated that high-performing students
produce substantially different HMM transition diagrams that tend to show
longer concentration span in quiz-taking and more active  forum
participation as compared with the average students.  These results show
the great potential of the proposed model for serving as a tool to help
humans discover knowledge about student behaviors.

@@

There are a number of interesting directions to further extend our work in
the future.  First, the proposed model is completely general and can thus
be easily applied to analyze the log of many other courses to enable deep
understanding of student behaviors as well as the correlations of such
behaviors and other variables such as grades. To realize these benefits, it
would be useful to develop a MOOC log analysis system based on the proposed
model so as to facilitate education research and help instructors improve
course design.

>>>>

We showed that such a model does, in fact, capture meaningful behavior
patterns and produces descriptions of these behavior patterns that are easy
to interpret. We argued that it is important to capture student behavior
patterns with more sophisticated models than simple discrete distributions
over actions to capture information present in bigrams of actions (or
larger sequences). By varying the number of latent states inferred, we
showed that the model is flexible and can capture patterns of differing
levels of specificity in this way.  Finally, we investigated whether we can
detect differences in student behavior patterns as they correlate with
course performance. Specifically, we demonstrated that high-performing
students produce substantially different HMM transition diagrams that tend
to show longer concentration span in quiz-taking and more active  forum
participation as compared with the average students.  These results show
the great potential of the proposed model for serving as a tool to help
humans discover knowledge about student behaviors.

@@

There are a number of interesting directions to further extend our work in
the future.  First, the proposed model is completely general and can thus
be easily applied to analyze the log of many other courses to enable deep
understanding of student behaviors as well as the correlations of such
behaviors and other variables such as grades. To realize these benefits, it
would be useful to develop a MOOC log analysis system based on the proposed
model to facilitate education research and help instructors improve course
design.

### 2017-08-28 sec-intro.tex
Address copy-editing feedback.

How should we represent behavioral patterns, and what does it mean to
understand changes in student behavior with respect to these patterns?
These are still very open questions and are active areas of
research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
Shih:2010:EDM}. In this paper, we advocate for a particular representation
of student behavior patterns as well as \emph{behavior transitions} that we
believe is simultaneously interpretable but also amenable to unsupervised,
automated discovery via statistical means. Specifically, we choose to
visualize behavior patterns as labeled directed graphs where a node
represents a ``behavior state'' (such as watching a lecture video or
visiting a forum), a directed edge indicates a transition from one
behavior state to another, node sizes are proportional to steady-state
probabilities, and edge widths are proportional to the probability of
leaving a node following that edge. We can use this same representation for
visualizing both the student behavior patterns as well as the transitioning
behavior between them. In Figure~\ref{fig:motivating-example} we show a
hypothetical example of the kind of output our proposed representation
could convey.  Here we see two different behavioral patterns
(\ref{fig:motivating-example-0} and \ref{fig:motivating-example-1}) as well
as the \emph{transition behavior} between these two behavior patterns
(\ref{fig:motivating-example-trans}).
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state0.png}
    \caption{Behavior pattern 0\label{fig:motivating-example-0}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state1.png}
    \caption{Behavior pattern 1\label{fig:motivating-example-1}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/trans.png}
    \caption{Transitions between the two behavior patterns to the
    left.\label{fig:motivating-example-trans}}
  \end{subfigure}
  \caption{An idealized example of what our behavior representation could capture.}
  \label{fig:motivating-example}
\end{figure}
Such a visualized state-transition representation is very informative for
describing student behavior. Indeed, we could infer many things from even
such a simple example: The first might be that, when students are taking
quizzes, they tend to either use the forum or the videos for support, but
not both. They also tend to take quizzes in a sort of ``cycle'' pattern,
indicating perhaps that this course allows quiz re-takes. Finally, in
Figure~\ref{fig:motivating-example-trans} we could conclude that users tend
to change their quiz-taking behavior over time from one that is more
video-focused (pattern 0) to one that is more forum-focused (pattern 1).

Our goal in this paper is to design a model that can automatically capture
student behavior in this way via unsupervised learning methods applied to
large collections of click logs associated with MOOCs. We view our model
as a component of a system that enables collaboration between the machine
and a human instructor to extract knowledge from large collections of
MOOC data. Automatically extracting interpretable patterns from the
clickstream data associated with MOOCs is a necessary step in order for
instructors to identify the hidden knowledge in massive interaction
datasets. Without the availability of a suitable model for identifying
behavioral patterns, instructors are not empowered to use this available
data to improve their courses without expending extraordinary amounts of
manual effort (even with which the raw data can still be very hard to
interpret).

Our proposed model (as well as the proposed behavior representation) is
motivated by the following observations:
\begin{enumerate}
  \item Student behavior is complicated and cannot necessarily be captured
      sufficiently by rule-based methods such as those explored by
      \citet{Kizilcec:2013:LAK} and \citet{Davis:2016:EDM}. We instead
      propose to treat student behavior patterns as being characterized
      (represented) via a sequence of \emph{latent states}. This allows us
      to automatically capture patterns that we might not have been able to
      articulate clearly a priori via a series of rules, and also allows us
      to model the inherent uncertainty in assigning a student's behavior
      to a particular pattern or group.
  \item Student behavior can vary over time. Previous models that treat
      students as exhibiting only one behavioral pattern over
      time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
      student behavior dynamics in a course. We propose a latent space model
      with \emph{latent state transitions} to flexibly model the dynamics.
  \item Analysis of student behavior can and should be performed at varying
      levels of granularity. This requires us to aggregate data over time
      with \emph{different levels of resolution}; existing models tend to come
      with a particular assumption about the resolution of time they
      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We
      propose a model that is agnostic to the time resolution considered,
      allowing it to be applied at different levels of resolution more
      naturally.
\end{enumerate}

>>>>

How should we represent behavioral patterns, and what does it mean to
understand changes in student behavior with respect to these patterns?
These are still very open questions and are active areas of
research~\cite{Kizilcec:2013:LAK, Faucon:2016:EDM, Davis:2016:EDM,
Shih:2010:EDM}. In this paper, we advocate for a representation of student
behavior patterns as well as \emph{behavior transitions} that we believe is
simultaneously interpretable but also amenable to unsupervised, automated
discovery via statistical means. Specifically, we choose to visualize
behavior patterns as labeled directed graphs where a node represents a
``behavior state'' (such as watching a lecture video or visiting a forum),
a directed edge indicates a transition from one behavior state to another,
node sizes are proportional to steady-state probabilities, and edge widths
are proportional to the probability of leaving a node following that edge.
We can use this same representation for visualizing both the student
behavior patterns as well as the transitioning behavior between them. In
Figure~\ref{fig:motivating-example} we show a hypothetical example of the
kind of output our proposed representation could convey.  Here we see two
different behavioral patterns (\ref{fig:motivating-example-0} and
\ref{fig:motivating-example-1}) as well as the \emph{transition behavior}
between these two behavior patterns (\ref{fig:motivating-example-trans}).
\begin{figure}
  \centering
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state0.png}
    \caption{Behavior pattern 0\label{fig:motivating-example-0}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/state1.png}
    \caption{Behavior pattern 1\label{fig:motivating-example-1}}
  \end{subfigure}
  \begin{subfigure}[t]{0.30\textwidth}
    \includegraphics[width=\textwidth]{figures/example/trans.png}
    \caption{Transitions between the two behavior patterns to the
    left.\label{fig:motivating-example-trans}}
  \end{subfigure}
  \caption{An idealized example of what our behavior representation could capture.}
  \label{fig:motivating-example}
\end{figure}
Such a visualized state-transition representation is very informative for
describing student behavior. Indeed, we could infer many things from even
such a simple example: The first might be that, when students are taking
quizzes, they tend to either use the forum or the videos for support, but
not both. They also tend to take quizzes in a sort of ``cycle'' pattern,
indicating perhaps that this course allows quiz re-takes. Finally, in
Figure~\ref{fig:motivating-example-trans} we could conclude that users tend
to change their quiz-taking behavior over time from one that is more
video-focused (pattern 0) to one that is more forum-focused (pattern 1).

Our goal in this paper is to design a model that can automatically capture
student behavior in this way via unsupervised learning methods applied to
large collections of click logs associated with MOOCs. We view our model
as a component of a system that enables collaboration between the machine
and a human instructor to extract knowledge from large collections of
MOOC data. Automatically extracting interpretable patterns from the
clickstream data associated with MOOCs is a necessary step for instructors
to identify the hidden knowledge in massive interaction datasets. Without
the availability of a suitable model for identifying behavioral patterns,
instructors are not empowered to use this available data to improve their
courses without expending extraordinary amounts of manual effort (even with
which the raw data can still be very hard to interpret).

Our proposed model (as well as the proposed behavior representation) is
motivated by the following observations:
\begin{enumerate}
  \item Student behavior is complicated and cannot necessarily be captured
      sufficiently by rule-based methods such as those explored by
      \citet{Kizilcec:2013:LAK} and \citet{Davis:2016:EDM}. We instead
      propose to treat student behavior patterns as being characterized
      (represented) via a sequence of \emph{latent states}. This allows us
      to automatically capture patterns that we might not have been able to
      articulate clearly a priori via a series of rules, and allows us to
      model the inherent uncertainty in assigning a student's behavior to a
      pattern or group.
  \item Student behavior can vary over time. Previous models that treat
      students as exhibiting only one behavioral pattern over
      time~\cite{Faucon:2016:EDM} miss out on the opportunity to understand
      student behavior dynamics in a course. We propose a latent space model
      with \emph{latent state transitions} to flexibly model the dynamics.
  \item Analysis of student behavior can and should be performed at varying
      levels of granularity. This requires us to aggregate data over time
      with \emph{different levels of resolution}; existing models tend to come
      with an assumption about the resolution of time they
      consider~\cite{Faucon:2016:EDM, Kizilcec:2013:LAK, Shih:2010:EDM}. We
      propose a model that is agnostic to the time resolution considered,
      allowing it to be applied at different levels of resolution more
      naturally.
\end{enumerate}

### 2017-08-28 sec-limitations.tex
Address copy-editing feedback.

There are a few limitations of our model that are important to highlight.
First, there are specific technical limitations due to the statistical
nature of the model and the particular methodology we propose for fitting
our model parameters. Second, there are limitations in the kinds of
patterns our model is able to discover in its current formulation and the
ease with which instructors are able to extract knowledge from these
patterns. We discuss both of these lines below.

@@

One potential limitation is that the model is complex and has many parameters
in order to truly uncover the relevant patterns in the data. In order to
properly estimate these parameters at training time, a large amount of data
must be available to the training algorithm. The assumption that we have a
large amount of sequence data available for training on should generally
hold for most MOOC courses, but this assumption may be problematic if
attempting to apply our model to smaller online (or on-campus) courses.

Our model fits its parameters using maximum likelihood estimation using the
EM algorithm. The EM algorithm is a hill climbing algorithm that is
optimizing in a highly non-convex parameter space. Thus, it can only
guarantee that we reach a local maximum in
practice~\citep{Dempster:1977:JRSS}.  This may mean that the parameters
found for the model may not be the ``best'' parameters in a global sense,
which may lead to suboptimal latent state representations and transition
patterns. Empirically, however, we observed in our experiments that strong
patterns tend to always show up as long as the algorithm reaches a
reasonably good local maximum, and the differences of the results tend to
be related to weak ``unstable'' patterns which may not be reliable anyway.
Since it is far more important and useful to reveal the strong salient
patterns than weak unreliable ones for instructors, the problem might not
necessarily affect the utility of the approach so significantly.  A
commonly applied approach to address the problem of multiple local maxima
is to run the model multiple times with different starting points to allow
the model to explore a larger portion of the potential parameter space. One
can then compare the log-likelihood of the data between the models that
were started from different initialization points and select the one that
has the highest value. This still does not guarantee that we find a global
maximum, but it does help alleviate the potential for finding a
particularly bad local maximum.  In practice, we have found that our model
can be fit to the data fairly quickly on commodity hardware, so running it
multiple times to address this concern is not computationally unfeasible.

@@

As is the case with traditional hidden Markov models, it is often important
to smooth the underlying model's distributions to ensure that there is a
non-zero probability of generating the observations. We employed a simple
additive smoothing in our implementation with a small additive constant for
all of our transition matrices to avoid this problem.

@@

The proposed behavior representation is most suitable for representing
recurring behavior patterns, which presumably are most interesting
to extract from the data, but may not cover all the interesting patterns in the data.
It would be interesting to further explore other complementary models such as
 time series models, which may help capture non-recurring patterns.

Our proposed model is flexible in the patterns it can discover in two main
ways. The first is that the granularity of the patterns can be adjusted by
changing the segmentation strategy one uses to divide the user action
stream into discrete ``sessions.'' The other is the number of latent
states $K$ that are used to describe the segmented action sequences. One
drawback of these two lines of flexibility is that there is not necessarily
a clear answer as to the ``correct'' approach for both in any given
scenario. Varying the segmentation strategy changes the granularity of the
patterns that are required to be explained by the latent states, which will
change their meaning. Varying the number of latent states increases the
flexibility of the model, but also may result in latent states that are not
substantially different from the other latent states and/or latent states
with very low probability. This flexibility forces a user of our model to
make some assumptions about the patterns they wish to find (granularity,
diversity), and the model itself does not necessarily provide clear
guidance as to what the best approach is.

Furthermore, we have made an implicit assumption that the segmentation
strategy is applied as a pre-processing step (and is most obviously a
deterministic process). The proposed segmentation strategies in this paper
do not specifically allow for the transitioning between the different
latent behavior patterns to occur over different windows of time for
different users: we have made a strong assumption that transitions between
latent states only occur at action sequence boundaries. One can model how
much time a user stays in a given state in terms of the number of sessions
they remain there before transitioning, but it may be better to directly
model the amount of time we expect a user to stay in the given state
directly. In other words, a more powerful model might be one in which the
segmentation and the latent behavior pattern discovery are jointly modeled
in a single probabilistic framework rather than being completely separate
pieces as we investigate in this work.

>>>>

There are a few limitations of our model that are important to highlight.
First, there are specific technical limitations due to the statistical
nature of the model and the particular methodology we propose for fitting
our model parameters. Second, there are limitations in the kinds of
patterns our model is able to discover in its current formulation and the
ease with which instructors are able to extract knowledge from these
patterns. We discuss both lines below.

@@

One potential limitation is that the model is complex and has many parameters
in order to truly uncover the relevant patterns in the data. To properly
estimate these parameters at training time, a large amount of data must be
available to the training algorithm. The assumption that we have a large
amount of sequence data available for training on should hold for most MOOC
courses, but this assumption may be problematic if attempting to apply our
model to smaller online (or on-campus) courses.

Our model fits its parameters using maximum likelihood estimation using the
EM algorithm. The EM algorithm is a hill climbing algorithm that is
optimizing in a highly non-convex parameter space. Thus, it can only
guarantee that we reach a local maximum in
practice~\citep{Dempster:1977:JRSS}.  This may mean that the parameters
found for the model may not be the ``best'' parameters in a global sense,
which may lead to suboptimal latent state representations and transition
patterns. Empirically, however, we observed in our experiments that strong
patterns tend to always show up if algorithm reaches a reasonably good
local maximum, and the differences of the results tend to be related to
weak ``unstable'' patterns which may not be reliable anyway.  Since it is
far more important and useful to reveal the strong salient patterns than
weak unreliable ones for instructors, the problem might not necessarily
affect the utility of the approach so significantly.  A commonly applied
approach to address the problem of multiple local maxima is to run the
model multiple times with different starting points to allow the model to
explore a larger portion of the parameter space. One can then compare the
log-likelihood of the data between the models that were started from
different initialization points and select the one that has the highest
value. This still does not guarantee that we find a global maximum, but it
does help alleviate the potential for finding a particularly bad local
maximum. In practice, we have found that our model can be fit to the data
quickly on commodity hardware, so running it multiple times to address this
concern is not computationally unfeasible.

@@

As is the case with traditional hidden Markov models, it is often important
to smooth the underlying model's distributions to ensure that there is a
non-zero probability of generating the observations. We employed a simple
additive smoothing in our implementation with a small additive constant
($10^{-6}$) for all our transition matrices to avoid this problem.

@@

The proposed behavior representation is most suitable for representing
recurring behavior patterns, which presumably are most interesting to
extract from the data, but may not cover all the interesting patterns in
the data. It would be interesting to explore other complementary models
such as time series models, which may help capture non-recurring patterns.

Our proposed model is flexible in the patterns it can discover in two main
ways. The first is that the granularity of the patterns can be adjusted by
changing the segmentation strategy one uses to divide the user action
stream into discrete ``sessions.'' The other is the number of latent
states $K$ that are used to describe the segmented action sequences. One
drawback of these two lines of flexibility is that there is not necessarily
a clear answer as to the ``correct'' approach for both in any given
scenario. Varying the segmentation strategy changes the granularity of the
patterns the latent states explain, which will change their meaning.
Varying the number of latent states increases the flexibility of the model,
but also may result in latent states that are not substantially different
from the other latent states and/or latent states with very low
probability. This flexibility forces a user of our model to make some
assumptions about the patterns they wish to find (granularity, diversity),
and the model itself does not necessarily provide clear guidance as to what
the best approach is.

Furthermore, we have made an implicit assumption that the segmentation
strategy is applied as a pre-processing step (and is most obviously a
deterministic process). The proposed segmentation strategies in this paper
do not specifically allow for the transitioning between the different
latent behavior patterns to occur over different windows of time for
different users: we have made a strong assumption that transitions between
latent states only occur at action sequence boundaries. One can model how
much time a user stays in each state in terms of the number of sessions
they remain there before transitioning, but it may be better to directly
model the amount of time we expect a user to stay in the given state.  In
other words, a more powerful model might be one in which the segmentation
and the latent behavior pattern discovery are jointly modeled in a single
probabilistic framework rather than being separate pieces as we investigate
in this work.

### 2017-08-28 sec-model.tex
Address copy-editing feedback.

An HMM is a specific probabilistic generative model with a ``built-in''
state transition system that would control the data to be generated by the
model, thus it is especially suitable for modeling sequence
data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would
be in one of $k$ states $U=\{u_1,\ldots,u_k\}$, and at the next moment, the
HMM would move to another state stochastically according to a transition
matrix that specifies the probability $p(u_i \mid u_j)$ of going to state
$u_i$ when the HMM is currently in state $u_j$. When the HMM is in state
$u$, the HMM can generate an observable data point $x$ according to an
output probabilistic model $p(x \mid u)$. Thus if we ``run'' an HMM for $N$
time points denoted by $t=1,\ldots,N$, the HMM could ``generate'' a sequence
of observations $x_1 \ldots x_N$, where each $x_i$ is an output symbol by
going through a sequence of \emph{hidden states} $w_1 \ldots w_N$ where
$w_i \in U$ is a state. The association of such a latent sequence of state
transitions with the observed symbols makes it possible to use the HMM to
``decode'' the latent behavior of students behind the surface behavior we
directly observe in the log data, allowing for understanding student
behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the
actual behavior of a student. We may say that students transition through
different ``task states'' (or ``behavior states'')  in the process of
study.  One such task state may be to learn about a topic by mostly
watching lecture videos, another task state may be to work on quizzes, and
yet another may be to participate in forum discussions. While in each of
these different states, the student would tend to exhibit different surface
``micro'' behaviors. For example, in the lecture study state, the student
would tend to have many video-watching related behaviors and occasionally
forum activities, while in the quiz-taking state (in order to pass each
module), the student would tend to show many quiz-related ``micro''
activities as well as asking questions or checking discussions on the
forum. Note that due to the complexity of the student behavior, it is very
difficult to accurately \emph{prescribe} the specific surface ``micro''
behavior patterns for each state in advance, especially without  prior
knowledge about the students. For example, forum activities are likely
interleaved with other activities   in every task state and the
interleaving pattern can be somewhat irregular with potentially many
variations.  The major motivations for using an HMM are that
\begin{enumerate*}[label=(\arabic*)]
  \item it uses a probabilistic model (the output probability distribution
    $p(x \mid u)$ conditioned on each state) to directly capture the
    inevitable uncertainty in the association of surface ``micro''
    activities with their corresponding latent task/behavior state, which
    is often our main target to discover and characterize, and
  \item it does not make any assumption about which latent task/behavior
    state must be associated with which observed activities or how a
    student would move from one state to another, but instead allows our
    data to ``tell'' us what kind of associations are most likely, what
    kind of transitions are most probable, and which states tend to be more
    long-lasting for any particular set of students.
\end{enumerate*}

However, if we use an ordinary HMM to analyze our data, we would treat each
observed ``micro'' activity (such as video watching or forum post reading)
as an output symbol, and thus the output distribution $p(x \mid u)$ for each
discovered latent state would be a simple distribution over all kinds of
observable micro activities recorded in our log data (e.g., 50\% lecture
watching, 8\% quiz taking, 7\% quiz submission, 2\% course wiki reading
\ldots). While such a distribution is meaningful and can already help us
interpret the corresponding latent state, it only gives us a rather
superficial characterization of student behavior.

@@

Adding this second layer would allow us to characterize a latent task
state in much more detail, as it would reveal not only what activities are
most common to a task state, but also the transition patterns between these
``micro'' activities (e.g., it can reveal frequent back-and-forth
transitions between quiz-taking and quiz-submission, which would suggest a
concentrated period of taking quizzes). Combining this ``surface'' Markov
model over the ``micro'' actions with the ``deep'' hidden Markov model over
the latent task states gives us a fairly general and powerful two-layer HMM
(2L-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
states and their transitions as well as the corresponding ``micro''
activity transition patterns associated with each latent state to
facilitate interpretation and analysis of the discovered latent state
patterns.

@@

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture,'' ``taking quiz,'' or ``viewing
forum.'' For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{O}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{o} \in
\mathbf{O}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_T)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper, we chose to denote the end of a sequence as occurring when no further
actions occur within a 10-hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and thus different
meanings behind the patterns that the 2L-HMM will extract. The flexibility
of using different segmentation strategies is intentional as it allows a
user to adjust the segmentation as needed so as to obtain patterns at
different granularity levels.

>>>>

An HMM is a specific probabilistic generative model with a ``built-in''
state transition system that would control the data to be generated by the
model, thus it is especially suitable for modeling sequence
data~\cite{Rabiner:1990:RSR, Huang:1990:HMM}. At any moment, the HMM would
be in one of $k$ states $U=\{u_1,\ldots,u_k\}$, and at the next moment, the
HMM would move to another state stochastically according to a transition
matrix that specifies the probability $p(u_i \mid u_j)$ of going to state
$u_i$ when the HMM is currently in state $u_j$. When the HMM is in state
$u$, the HMM can generate an observable data point $x$ according to an
output probabilistic model $p(x \mid u)$. Thus, if we ``run'' an HMM for $N$
time points denoted by $t=1,\ldots,N$, the HMM could ``generate'' a sequence
of observations $x_1 \ldots x_N$, where each $x_i$ is an output symbol by
going through a sequence of \emph{hidden states} $w_1 \ldots w_N$ where
$w_i \in U$ is a state. The association of such a latent sequence of state
transitions with the observed symbols makes it possible to use the HMM to
``decode'' the latent behavior of students behind the surface behavior we
directly observe in the log data, allowing for understanding student
behavior more deeply than a model with no latent state variables.

In many ways, the generation process behind an HMM is meant to simulate the
actual behavior of a student. We may say that students transition through
different ``task states'' (or ``behavior states'')  in the process of
study.  One such task state may be to learn about a topic by mostly
watching lecture videos, another task state may be to work on quizzes, and
yet another may be to participate in forum discussions. While in each of
these different states, the student would tend to exhibit different surface
``micro'' behaviors. For example, in the lecture study state, the student
would tend to have many video-watching related behaviors and occasionally
forum activities, while in the quiz-taking state (to pass each module), the
student would tend to show many quiz-related ``micro'' activities as well
as asking questions or checking discussions on the forum. Note that due to
the complexity of the student behavior, it is very difficult to accurately
\emph{prescribe} the specific surface ``micro'' behavior patterns for each
state in advance, especially without  prior knowledge about the students.
For example, forum activities are likely interleaved with other activities
in every task state and the interleaving pattern can be somewhat irregular
with potentially many variations. The major motivations for using an HMM
are that
\begin{enumerate*}[label=(\arabic*)]
  \item it uses a probabilistic model (the output probability distribution
    $p(x \mid u)$ conditioned on each state) to directly capture the
    inevitable uncertainty in the association of surface ``micro''
    activities with their corresponding latent task/behavior state, which
    is often our main target to discover and characterize, and
  \item it does not make any assumption about which latent task/behavior
    state must be associated with which observed activities or how a
    student would move from one state to another, but instead allows our
    data to ``tell'' us what kind of associations are most likely, what
    kind of transitions are most probable, and which states tend to be more
    long-lasting for any set of students.
\end{enumerate*}

However, if we were to use an ordinary HMM to analyze our data, we would
treat each observed ``micro'' activity (such as video watching or forum
post reading) as an output symbol, and thus the output distribution $p(x
\mid u)$ for each discovered latent state would be a simple distribution
over all kinds of observable micro activities recorded in our log data
(e.g., 50\% lecture watching, 8\% quiz taking, 7\% quiz submission, 2\%
course wiki reading, etc.). While such a distribution is meaningful and
can already help us interpret the corresponding latent state, it only gives
us a rather superficial characterization of student behavior.

@@

Adding this second layer would allow us to characterize a latent task
state in much more detail, as it would reveal not only what activities are
most common to a task state, but also the transition patterns between these
``micro'' activities (e.g., it can reveal frequent back-and-forth
transitions between quiz-taking and quiz-submission, which would suggest a
concentrated period of taking quizzes). Combining this ``surface'' Markov
model over the ``micro'' actions with the ``deep'' hidden Markov model over
the latent task states gives us a general and powerful two-layer HMM
(2L-HMM) that can simultaneously learn ``deeply'' the latent task/behavior
states and their transitions as well as the corresponding ``micro''
activity transition patterns associated with each latent state to
facilitate interpretation and analysis of the discovered latent state
patterns.

@@

Given a MOOC log, we can define a set $\mathbf{A}$ of actions that a
student can take at any given time. For example, an action $a \in
\mathbf{A}$ might be ``viewing lecture,'' ``taking quiz,'' or ``viewing
forum.'' For each student in the course $\ell \in \mathbf{L}$, we then
extract a list of action sequences $\mathbf{O}_\ell$ that he or she
produced as observed in the log, where each sequence $\mathbf{o} \in
\mathbf{O}_\ell$ is itself a list of actions $(a_1, a_2, \ldots, a_T)$ with
each $a_i \in \mathbf{A}$.  Each sequence can be divided flexibly; in this
paper, we chose to denote the end of a sequence as occurring when no further
actions occur within a 10-hour window of time (and thus our sequences
roughly correspond to one day's worth of activity). This decision reflects
our desire to uncover latent state transition behavior at the granularity
of day-to-day behavior. A different segmentation strategy could be used to
uncover hour-by-hour behavior or week-by-week behavior, and this depends
entirely on the desired time resolution one wishes to be exhibited in the
transitions between the latent states. Different segmentation strategies
will result in different underlying training data, and thus different
meanings behind the patterns that the 2L-HMM will extract. The flexibility
of using different segmentation strategies is intentional as it allows a
user to adjust the segmentation as needed to obtain patterns at different
granularity levels.
